Enhancing Adversarial Robustness for Deep Metric Learning

### 2021-10-21 appendix.tex
tex
remove deprecated tabs
appendix

@@

@@

\end{itemize}

@@

\input{direct-amd.tex}

\input{sort-hardness.tex}

@@

\input{amdsemi.tex}

>>>>

\section{Complete Table for Sec~4.1 Hardness Selection}

@@

	\item [TODO] arxiv 2105.04906 for collapse issue.

	\item [?] Can we learn more from the third and fourth standard moments?
		Namely skewness and kurtosis.

	\item [M] manuscript notes on SGA.

	\item [\cmark] We don't have to mix HM(R,M) and HM(S,M).

	\item [Q] Is ACT equivalent to $HM(H_\mathsf{S},0)$.

@@

	\item [T] Revise and incorporate the state-of-the-art adversarial training
		acceleration method (\ie, Free Adversarial Training~\cite{freeat},
		which is designed for classification) into adversarial training of 
		deep metric learning models. It will greatly improve the efficiency
		of adversarial training.
		%
	\item [T] Benchmark of existing metric learning loss functions under
		adversarial training scenario, and analyze their different
		characteristics for future reference.
		%
	\item [T] Explore the possibility of proposing a new metric learning loss
		function oriented for adversarial training, starting from scratch or
		from an existing metric learning loss.
		%
\end{itemize}

@@

@@

### 2021-10-21 robdml.tex
tex

Given a set of data points, a metric gives a distance value between each pair
of them.
%
Deep Metric Learning (DML) aims to learn such a metric between two inputs (\eg,
images) leveraging the representational power of deep neural networks.
%
DML has been extensively studied~\cite{revisiting}, and has a
wide range of applications such
as image retrieval~\cite{imagesim2} and face recognition~\cite{facenet,domainface}.

Despite the improvements of this field thanks to the advancement of deep neural
networks, recent research works suggest that DML models are vulnerable to
adversarial attacks, where an imperceptible perturbation could incur unexpected
or covertly manipulated results~\cite{advrank,advorder}.
%
Such vulnerability poses safety, security, and fairness concerns in the
applications of DML.
%
For example, impersonation or recognition evation are possible for DML-based
face-identification system.
%
To counter the attacks (or reducing the vulnerability), it is important to
design defense methods to improve the adversarial robustness of DML models.

@@

% existing methods & problem
Several defense methods for DML are proposed in the
literature~\cite{advrank,robrank}, inspired by the Madry's adversarial training
method~\cite{madry}, as it is kown as one of the most effective defense methods
for deep neural network classifiers.
%
However, it has been noted that
%
(1) the robustness level achieved by the existing methods is still insufficient
to counter the attacks;
%
(2) the direct adaptation of Madry's min-max adversarial training~\cite{madry} will easily
lead to model collapse due to producing very hard sample triplets;
%
(3) the adversarial training procedure is very time-consuming compared to
the training process of a regular DML model, but existing defense methods
are incompatible to acceleration methods like Free Adversarial Training~\cite{freeat}.

@@

% contributions
In brief, our contribusions include:
%
\begin{enumerate}[noitemsep]
	%
	\item {\textit{Hardness Manipulation}} is proposed for the adversarial
		training of triplet-based deep metric learning models, which avoids
		model collapse in the typical min-max adversarial training setting.
		The proposed method achieves higher adversarial robustness compared to
		the state-of-the-art, and is more computationally efficient.
		%
		This method makes min-max training feaisble, and hence the incorporation
		into free adversarial training and boost learning efficiency.
		%
		(initial condition)
		%
	\item \textit{Gradual Adversary} is proposed to dynamically adjust the
		destination hardness less for Hardness Manipulation during the
		adversarial training process in order to balance the metric learning
		and adversarial learning.
		%
		(terminal condition)
		%
	\item Inter-ID structure constraint. ((aa~p) + (pp~a))/2
		(cross-id repelling)
		existing methods only care inter-class separation, but not
		intra-class (inter-id) separation, because we may get a 6-element
		set (a,a~,p,p~,n,n~) from the original triplet.
		%
	\item Revise and incorporate the state-of-the-art adversarial training
		acceleration method (\ie, Free Adversarial Training~\cite{freeat},
		which is designed for classification) into adversarial training of 
		deep metric learning models. It will greatly improve the efficiency
		of adversarial training.
		%
	\item Benchmark of existing metric learning loss functions under
		adversarial training scenario, and analyze their different
		characteristics for future reference.
		%
	\item Explore the possibility of proposing a new metric learning loss
		function oriented for adversarial training, starting from scratch or
		from an existing metric learning loss.
		%
\end{enumerate}

@@

Such deficiency in the late training process can be alleviated with a
pseudo-hardness function that boosts the destination hardness when the loss
value is low.
%
Denoting the loss value from the previous training iteration as $\ell_{t-1}$,
we first normalize it into $[0,1]$ as $\bar{\ell}_{t-1}=\min(u,\ell_{t-1})/u$,
where $u$ is a manually specified parameter.
%
Note, $u$ is much less than the theoretical upperbound of the loss function,
in order to avoid excessive hardness boost.
%
Then the expectation of destination hardness can be linearly shifted (boosted)
by a scaled constant by a ``hardness boosting'' pseudo-hardness
function, \ie,
%
\begin{equation}
	%
	g_\mathsf{B}(H_\mathsf{D};\xi) = H_\mathsf{D} + \xi \cdot (1-\bar{\ell}_{t-1}).
	%
\end{equation}
%
And the adversarial examples from $\text{HM}(H_\mathsf{S},g_\mathsf{B}(H_\mathsf{D}))$
can be used for training.

@@

However, while the destination hardness is boosted in the late phase of
training, the adversarial examples may still hamper the learning of a good
embedding space in the early phase.
%
At this point, we argue that learning from a strong adversary in the early
training phase (\ie, when the embedding space is not ``in-good-shape'') hampers the
generalization on benign examples.
%
In other words, the desired destination hardness should be (1) relatively weak
in the early training phase, and (2) relatively strong in the late training
phase.
%
We name the family of pseudo-hardness functions that fulfills the two
conditions as ``Graduate Adversary''.
%
Additionally, if we have a strong prior knowledge on ``how hard the destination
hardness should be'', the pseudo-hardness function can be independent from
any benign triplet.

@@

For instance, if we assume that the destination hardness should remain
in the hardness range of Semihard~\cite{facenet}, the following function
satisfies the two condictions of Gradual Adversary, dubbed
``Simple Gradual Adversary'' (SGA):
%
\begin{equation}
	%
	g_\mathsf{SGA}() = -\gamma \cdot \bar{\ell}_{t-1} ~ \in [-\gamma,0].
	%
\end{equation}
%
The design of this function will be empirically supported in the experiments
section.
%
The exploration of other possible pseudo-hardness functions is omitted to 
keep simplicity.

>>>>

Given a set of data points, a metric gives a distance value between each pair
of them.
%
Deep Metric Learning (DML) aims to learn such a metric between two inputs (\eg,
images) leveraging the representational power of deep neural networks.
%
As an extensively studied task~\cite{revisiting}, DML has a wide range of
applications such as image retrieval~\cite{imagesim2} and face
recognition~\cite{facenet,domainface}, as well as a wide impact in
the core idea of other areas such as self-supervised learning~\cite{dmlreality}.

Despite the improvements of this field thanks to the advancement of deep neural
networks, recent research works suggest that DML models are vulnerable to
adversarial attacks, where an imperceptible perturbation could incur unexpected
or covertly manipulated results~\cite{advrank,advorder}.
%
Such vulnerability raises safety, security, and fairness concerns in the
applications of DML.
%
For example, impersonation or recognition evation are possible for DML-based
face-identification system.
%
To counter the attacks (\ie, mitigating the vulnerability), it is important to
design defense methods to improve the adversarial robustness of DML models.

@@

% existing methods & problem
Several defense methods for DML are proposed in the
literature~\cite{advrank,robrank}, inspired by the Madry's adversarial training
method~\cite{madry}, as it is kown as one of the most effective defense methods
for deep neural network classifiers.
%
However, it has been noted that
%
(1) the robustness level achieved by the existing methods is still insufficient
to counter the attacks;
%
(2) the direct adaptation of Madry's min-max adversarial training~\cite{madry} will easily
lead to model collapse due to producing very hard sample triplets;
%
(3) the adversarial training procedure is very time-consuming compared to
the training process of a regular DML model, but existing defense methods
are incompatible to acceleration methods like Free Adversarial Training~\cite{freeat}.
\oo{rob, time, r1}.

@@

% contributions
In brief, our contribusions include:
%
\begin{enumerate}[noitemsep]
	%
	\item {\textit{Hardness Manipulation}} is proposed for the adversarial
		training of triplet-based deep metric learning models, which avoids
		model collapse in the typical min-max adversarial training setting.
		The proposed method achieves higher adversarial robustness compared to
		the state-of-the-art, and is more computationally efficient.
		%
		This method makes min-max training feaisble, and hence the incorporation
		into free adversarial training and boost learning efficiency.
		%
		(initial condition)
		%
	\item \textit{Gradual Adversary} is proposed to dynamically adjust the
		destination hardness less for Hardness Manipulation during the
		adversarial training process in order to balance the metric learning
		and adversarial learning.
		%
		(terminal condition)
		%
	\item Inter-ID structure constraint. ((aa~p) + (pp~a))/2
		(cross-id repelling)
		existing methods only care inter-class separation, but not
		intra-class (inter-id) separation, because we may get a 6-element
		set (a,a~,p,p~,n,n~) from the original triplet.
		%
\end{enumerate}

The proposed method overwhelmingly outperforms the state-of-the-art defense
in terms of robustness, training cost, as well as the penalty on benign
example retrieval.

@@

Such deficiency in the late training process can be alleviated with a
pseudo-hardness function that boosts the destination hardness when the loss
value is low.
%
Denoting the loss value from the previous training iteration as $\ell_{t-1}$,
we first normalize it into $[0,1]$ as $\bar{\ell}_{t-1}=\min(u,\ell_{t-1})/u$,
where $u$ is a manually specified parameter.
%
Note, $u$ is much less than the theoretical upperbound of the loss function,
in order to avoid excessive hardness boost.
%
Then the expectation of destination hardness can be linearly shifted (boosted)
by a scaled constant by a ``hardness boosting'' pseudo-hardness
function, \ie,
%
\begin{equation}
	%
	g_\mathsf{B}(H_\mathsf{D};\xi,\bar{\ell}_{t-1}) =
	H_\mathsf{D} + \xi \cdot (1-\bar{\ell}_{t-1}).
	%
\end{equation}
%
And the adversarial examples from $\text{HM}(H_\mathsf{S},g_\mathsf{B}(H_\mathsf{D}))$
can be used for training.

@@

However, while the destination hardness is boosted in the late phase of
training, the adversarial examples may still hamper the learning of a good
embedding space in the early phase.
%
At this point, we argue that learning from a strong adversary in the early
training phase (\ie, when the embedding space is not ``in-good-shape'') hampers the
generalization on benign examples.

% Gradual Adversary

In other words, the desired destination hardness should be (1) relatively weak
in the early training phase, and (2) relatively strong in the late training
phase.
%
We name the family of pseudo-hardness functions that fulfills the two
conditions as ``Graduate Adversary''.
%
Additionally, if we have a strong prior knowledge on ``how hard the destination
hardness should be'', the pseudo-hardness function can be independent from
any benign triplet.

@@

For instance, if we assume that the destination hardness should remain
in the hardness range of Semihard~\cite{facenet}, the following function
satisfies the two condictions of Gradual Adversary, dubbed
``Simple Gradual Adversary'' (SGA):
%
\begin{equation}
	%
	g_\mathsf{SGA}(\bar{\ell}_{t-1}) =
	-\gamma \cdot \bar{\ell}_{t-1} ~ \in [-\gamma,0].
	%
\end{equation}
%
And the adversarial examples from $\text{HM}(H_\mathsf{S},g_\mathsf{SGA})$
can be used for adversarial training.
%
The design of this function will be empirically supported in the experiments
section.
%
The exploration of other possible pseudo-hardness functions is omitted 
for simplicity.
